An indexed construction of semi-simplicial and semi-cubical types

### 2022-05-17 paper/paper.tex
paper: fix error in universe letter in U
paper: drop universe letter on X in abbrev-tabs 3,4
paper: many little corrections to abbrev-tabs
paper: s/ETT/HSet in tables
paper: remove trailing comma in restr/coh with no arg

% Some lua magic to join a variable number of arguments with commas
% Comma-separeated optional arguments
\begin{luacode*}
  function remove_empty(tbl)
    for i, v in ipairs(tbl) do
        if v == "" then
          table.remove(tbl, i)
          return remove_empty(tbl)
        end
    end
    return tbl
  end
  function join_with_comma(tbl)
    return table.concat(remove_empty(tbl), ",")
  end
\end{luacode*}

\newcommandx{\joinargs}[5][1=,2=,3=,4=,5=]{\directlua{tex.sprint(join_with_comma({"\luaescapestring{#1}";"\luaescapestring{#2}";"\luaescapestring{#3}";"\luaescapestring{#4}";"\luaescapestring{#5}"}))}}

@@

% Definitional and extensional equality
\newcommand{\defeq}{\ensuremath{\triangleq}}
\newcommand{\eqett}{\ensuremath{\equiv_{\mathit{ETT}}}}
\newcommand{\reflett}{\ensuremath{\mathsf{refl}_{\mathit{ETT}}}}

% Truncated sets, otherwise referred to as X
\newcommand{\Xp}[1]{\ensuremath{\mathsf{\color{carolina}{X}}_{#1}}}
\newcommand{\X}[2]{\ensuremath{\mathsf{\color{carolina}{X}}_{#1}^{<#2}}}
\newcommand{\Xcomp}[2]{\ensuremath{\mathsf{\color{carolina}{X}}_{#1}^{=#2}}}
\newcommand{\Xfrom}[2]{\ensuremath{\mathsf{\color{carolina}{X}}_{#1}^{\geq#2}}}

% Frame, layer, and filler: the general macro
\newcommandx{\prim}[5][2=,3=,4=,5=]{\ensuremath{\mathsf{\color{indian-yellow}{#1}}_{#2}^{\joinargs[#3][#4][#5]}}}

% Restriction for frame, layer, and filler: the general macro
\newcommandx{\restr}[7][2=,3=,4=,5=,6=,7=]{\ensuremath{\mathsf{\color{russian-green}{restr}}_{\mathsf{\color{indian-yellow}{#1}}, \joinargs[#2][#3][#4]}^{\joinargs[#5][#6][#7]}}}

% Coherence conditions for frame, layer, and filler: the general macro
\newcommandx{\coh}[9][2=,3=,4=,5=,6=,7=,8=,9=]{\ensuremath{\mathsf{\color{chestnut}{coh}}_{\mathsf{\color{indian-yellow}{#1}}, \joinargs[#2][#3][#4][#5][#6]}^{\joinargs[#7][#8][#9]}}}

@@

\begin{document}
\thispagestyle{empty}
\maketitle
\begin{abstract}
  A model of \emph{homotopy type theory}, that formalizes \emph{truncated} $n$-types in dependent type theory. The accompanying artifact is an axiomless Coq formalization~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to truncated $n$-types; if one were to instantiate our development for a particular type (say, augmented simplicial types), this requirement vanishes.} that uses sophisticated rewriting techniques and pushes the boundary of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

@@

Our contribution is to model the core of semi-cubical sets that naturally generalizes to augmented simplicial sets in type theory, and we call our generalization \emph{truncated $n$-sets}. In contrast to previous efforts that use a fibered representation, we use an indexed representation, that can be derived from the fibered one using a \emph{iterated reverse-Grothendieck} construction. Our work is, in principle axiom-free, although the accompanying artifact uses functional extensionality in order to generalize our work to $n$-truncated sets.

@@

\subsection{\texorpdfstring{\DeltaHat}{Delta sets}}
As a first step towards formalizing simplicial sets, we restrict outselves to representing \emph{augmented simplicial sets}, which are described roughly as delta-sets in mathematical literature, which forgets about the degeneracies. We now proceed to describe \DeltaHat, which is identical to \Simplex{}, but for the fact that the maps $\sq{m} \rightarrow \sq{n}$ are \emph{strictly} order-preserving. The degeneracies then vanish, and we're left with the task of defining gluing conditions solely on the basis of face maps. In mathematical literature, there is little interest in studying \DeltaHat, since maps between delta sets are not well-defined, and it is therefore not possible to construct a category of delta sets.

@@

\begin{theorem}
  $\CSet$ does not admit degeneracies.
\end{theorem}

\begin{proof}
  The reader is advised to refer to \cite{Antolini00} for the proof.
\end{proof}

@@

% Now, we write the type theory bit, and input the abbreviated tables
% Frame, layer, and filler
\newcommandx{\framep}[4][1=,2=,3=,4=]{\prim{frame}[][#2][#3][#4]}
\newcommandx{\layer}[4][1=,2=,3=,4=]{\prim{layer}[][#2][#3][#4]}
\newcommandx{\filler}[4][1=,2=,3=,4=]{\prim{filler}[][#2][#3][#4]}
\newcommandx{\fullframe}[2][1=,2=]{\prim{fullframe}[][#2][][]}
% Restriction for frame, layer, and filler
\newcommandx{\restrf}[6][1=,2=,3=,4=,5=,6=]{\restr{frame}[][#2][][#4][#5][#6]}
\newcommandx{\restrl}[6][1=,2=,3=,4=,5=,6=]{\restr{layer}[][#2][][#4][#5][#6]}
\newcommandx{\restrc}[6][1=,2=,3=,4=,5=,6=]{\restr{filler}[][#2][][#4][#5][#6]}

% Coherence conditions for frame, layer, and filler
\newcommandx{\cohf}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{frame}[][#2][#3][][][][][#8]}
\newcommandx{\cohl}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{layer}[][#2][#3][][][][][#8]}
\newcommandx{\cohc}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{filler}[][#2][#3][][][][][#8]}

@@

\input{tab-core.tex}
\input{tab-frames.tex}
\input{tab-faces.tex}

% Restriction for frame, layer, and filler
\renewcommandx{\restrf}[6][1=,2=,3=,4=,5=,6=]{\restr{frame}[][#2][#3][#4][#5][#6]}
\renewcommandx{\restrl}[6][1=,2=,3=,4=,5=,6=]{\restr{layer}[][#2][#3][#4][#5][#6]}
\renewcommandx{\restrc}[6][1=,2=,3=,4=,5=,6=]{\restr{filler}[][#2][#3][#4][#5][#6]}

% Coherence conditions for frame, layer, and filler
\renewcommandx{\cohf}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{frame}[][][][#4][#5][][#7][#8]}
\renewcommandx{\cohl}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{layer}[][][][#4][#5][][#7][#8]}
\renewcommandx{\cohc}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{filler}[][][][#4][#5][][#7][#8]}

\input{tab-coh.tex}
\input{tab-coind.tex}

@@

\newpage
\appendix
\section{Tables in full detail}
% Frame, layer, and filler
\renewcommandx{\framep}[4][1=,2=,3=,4=]{\prim{frame}[#1][#2][#3][#4]}
\renewcommandx{\layer}[4][1=,2=,3=,4=]{\prim{layer}[#1][#2][#3][#4]}
\renewcommandx{\filler}[4][1=,2=,3=,4=]{\prim{filler}[#1][#2][#3][#4]}
\renewcommandx{\fullframe}[2][1=,2=]{\prim{fullframe}[#1][#2][][]}

@@

\newpage

\begin{thebibliography}{10}
  \bibitem[Bez]{Bezem14}
  Bezem, M., Coquand, T., \& Huber, S. (2014, July). A model of type theory in cubical sets. \textit{In 19th International Conference on Types for Proofs and Programs (TYPES 2013)} (Vol. 26, pp. 107-128). Wadern, Germany: Schloss Dagstuhl–Leibniz Zentrum fuer Informatik.

>>>>

% Some lua magic to join a variable number of arguments with commas
% Comma-separeated optional arguments
\begin{luacode*}
  function remove_empty(tbl)
    for i, v in ipairs(tbl) do
        if v == "" then
          table.remove(tbl, i)
          return remove_empty(tbl)
        end
    end
    return tbl
  end
  function join_with_comma(prefix, tbl)
    str = table.concat(remove_empty(tbl), ",")
    if str == "" then
      return str
    end
    return prefix .. str
  end
\end{luacode*}

\newcommandx{\joinargs}[6][2=,3=,4=,5=,6=]{\directlua{tex.sprint(join_with_comma("#1", {"\luaescapestring{#2}";"\luaescapestring{#3}";"\luaescapestring{#4}";"\luaescapestring{#5}";"\luaescapestring{#6}"}))}}

@@

% Definitional and extensional equality
\newcommand{\defeq}{\ensuremath{\triangleq}}
\newcommand{\eqHSet}{\ensuremath{\equiv_{\mathit{HSet}}}}
\newcommand{\reflHSet}{\ensuremath{\mathsf{refl}_{\mathit{HSet}}}}

% Truncated sets, otherwise referred to as X: the general macro
\newcommandx{\X}[2][1=,2=]{\ensuremath{\mathsf{\color{carolina}{X}}_{#1}^{#2}}}

% Frame, layer, and filler: the general macro
\newcommandx{\prim}[5][2=,3=,4=,5=]{\ensuremath{\mathsf{\color{indian-yellow}{#1}}_{#2}^{\joinargs{}[#3][#4][#5]}}}

% Restriction for frame, layer, and filler: the general macro
\newcommandx{\restr}[7][2=,3=,4=,5=,6=,7=]{\ensuremath{\mathsf{\color{russian-green}{restr}}_{\mathsf{\color{indian-yellow}{#1}}\joinargs{,}[#2][#3][#4]}^{\joinargs{}[#5][#6][#7]}}}

% Coherence conditions for frame, layer, and filler: the general macro
\newcommandx{\coh}[9][2=,3=,4=,5=,6=,7=,8=,9=]{\ensuremath{\mathsf{\color{chestnut}{coh}}_{\mathsf{\color{indian-yellow}{#1}}\joinargs{,}[#2][#3][#4][#5][#6]}^{\joinargs{}[#7][#8][#9]}}}

@@

\begin{document}
\thispagestyle{empty}
\maketitle
\begin{abstract}
  A model of \emph{homotopy type theory}, that formalizes \emph{truncated} $n$-types in dependent type theory. The accompanying artifact is an axiomless Coq formalization~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to truncated $n$-types; if one were to instantiate our development for a particular type (say, semi-simplicial types), this requirement vanishes.} that uses sophisticated rewriting techniques and pushes the boundary of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

@@

Our contribution is to model the core of semi-cubical sets that naturally generalizes to semi-simplicial sets in type theory, and we call our generalization \emph{truncated $n$-sets}. In contrast to previous efforts that use a fibered representation, we use an indexed representation, that can be derived from the fibered one using a \emph{iterated reverse-Grothendieck} construction. Our work is, in principle axiom-free, although the accompanying artifact uses functional extensionality in order to generalize our work to $n$-truncated sets.

@@

\subsection{\texorpdfstring{\DeltaHat}{Delta sets}}
The first step towards formalizing simplicial sets is to forget about the degeneracies, and we restrict ourselves to representing \emph{augmented simplicial sets}, otherwise known as delta-sets in mathematical literature. We now proceed to describe \DeltaHat, which is identical to \Simplex{}, but for the fact that the maps $\sq{m} \rightarrow \sq{n}$ are \emph{strictly} order-preserving. The degeneracies then vanish, and we're left with the task of defining gluing conditions solely on the basis of face maps. In mathematical literature, there is little interest in studying \DeltaHat, since maps between delta sets are not well-defined, and it is therefore not possible to construct a category of delta sets.

@@

\begin{theorem}
  $\CSet$ does not admit degeneracies.
  \begin{proof}
    The reader is advised to refer to \cite{Antolini00} for the proof.
  \end{proof}
\end{theorem}

@@

% Now, we write the type theory bit, and input the abbreviated tables

% Truncated sets, otherwise referred to as X
% Keep all arguments
\newcommandx{\Xp}[1]{\X[#1][]}
\newcommandx{\Xto}[2]{\X[#1][<#2]}
\newcommandx{\Xcomp}[2]{\X[#1][=#2]}
\newcommandx{\Xfrom}[2]{\X[#1][\geq#2]}

% Frame, layer, and filler
% Drop the universe letter
\newcommandx{\framep}[4][1=,2=,3=,4=]{\prim{frame}[][#2][#3][#4]}
\newcommandx{\layer}[4][1=,2=,3=,4=]{\prim{layer}[][#2][#3][#4]}
\newcommandx{\filler}[4][1=,2=,3=,4=]{\prim{filler}[][#2][#3][#4]}

% Restriction for frame, layer, and filler
% Drop the unviverse letter and q
\newcommandx{\restrf}[6][1=,2=,3=,4=,5=,6=]{\restr{frame}[][#2][][#4][#5][#6]}
\newcommandx{\restrl}[6][1=,2=,3=,4=,5=,6=]{\restr{layer}[][#2][][#4][#5][#6]}
\newcommandx{\restrc}[6][1=,2=,3=,4=,5=,6=]{\restr{filler}[][#2][][#4][#5][#6]}

% Coherence conditions for frame, layer, and filler
% Keep only ε, ω; n, p; proposition on n and p
\newcommandx{\cohf}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{frame}[][#2][#3][][][#6][#7][#8]}
\newcommandx{\cohl}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{layer}[][#2][#3][][][#6][#7][#8]}
\newcommandx{\cohc}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{filler}[][#2][#3][][][#6][#7][#8]}

@@

% For just the first table, keep the universe letter in fullframe
\newcommandx{\fullframe}[2][1=,2=]{\prim{fullframe}[#1][#2][][]}
\input{tab-core.tex}

% Drop the universe letter in fullframe for all remaining tables
\renewcommandx{\fullframe}[2][1=,2=]{\prim{fullframe}[][#2][][]}
\input{tab-frames.tex}

% Truncated sets, otherwise referred to as X
% Drop the universe letter for the third and fourth tables
\renewcommandx{\Xto}[2]{\X[][<#2]}
\renewcommandx{\Xcomp}[2]{\X[][=#2]}
\renewcommandx{\Xfrom}[2]{\X[][\geq#2]}
\input{tab-faces.tex}

% Restriction for frame, layer, and filler
% Drop just the universe letter for the coh table
\renewcommandx{\restrf}[6][1=,2=,3=,4=,5=,6=]{\restr{frame}[][#2][#3][#4][#5][#6]}
\renewcommandx{\restrl}[6][1=,2=,3=,4=,5=,6=]{\restr{layer}[][#2][#3][#4][#5][#6]}
\renewcommandx{\restrc}[6][1=,2=,3=,4=,5=,6=]{\restr{filler}[][#2][#3][#4][#5][#6]}

% Frame, layer, and filler
% Drop all arguments for coh table; they're not present in coind table.
\renewcommandx{\framep}[4][1=,2=,3=,4=]{\prim{frame}[][][][]}
\renewcommandx{\layer}[4][1=,2=,3=,4=]{\prim{layer}[][][][]}
\renewcommandx{\filler}[4][1=,2=,3=,4=]{\prim{filler}[][][][]}

% Coherence conditions for frame, layer, and filler
% Drop universe letter, ε, and ω
\renewcommandx{\cohf}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{frame}[][][][#4][#5][#6][#7][#8]}
\renewcommandx{\cohl}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{layer}[][][][#4][#5][#6][#7][#8]}
\renewcommandx{\cohc}[8][1=,2=,3=,4=,5=,6=,7=,8=]{\coh{filler}[][][][#4][#5][#6][#7][#8]}
\input{tab-coh.tex}

% Truncated sets, otherwise referred to as X
% Keep universe letter for last table
\renewcommandx{\Xto}[2]{\X[#1][<#2]}
\renewcommandx{\Xcomp}[2]{\X[#1][=#2]}
\renewcommandx{\Xfrom}[2]{\X[#1][\geq#2]}

\input{tab-coind.tex}

@@

\newpage
\appendix
\section{Tables in full detail}
% Keep all arguments for all macros
% Frame, layer, and filler
\renewcommandx{\framep}[4][1=,2=,3=,4=]{\prim{frame}[#1][#2][#3][#4]}
\renewcommandx{\layer}[4][1=,2=,3=,4=]{\prim{layer}[#1][#2][#3][#4]}
\renewcommandx{\filler}[4][1=,2=,3=,4=]{\prim{filler}[#1][#2][#3][#4]}
\renewcommandx{\fullframe}[2][1=,2=]{\prim{fullframe}[#1][#2][][]}

@@

\newpage
\begin{thebibliography}{10}
  \bibitem[Bez]{Bezem14}
  Bezem, M., Coquand, T., \& Huber, S. (2014, July). A model of type theory in cubical sets. \textit{In 19th International Conference on Types for Proofs and Programs (TYPES 2013)} (Vol. 26, pp. 107-128). Wadern, Germany: Schloss Dagstuhl–Leibniz Zentrum fuer Informatik.
